Comparing \pkg{binsegRcpp} with other implementations of binary segmentation for change-point detection

### 2025-05-08 article.Rnw
intro

%% - \Abstract{} almost as usual
\Abstract{
  Binary segmentation is a classic algorithm for detecting
  change-points in sequential data. In theory, using a simple loss
  function like Normal or Poisson, binary segmentation should be
  extremely fast for $N$ data and $K$ segments: asymptotically $O(N K)$ time in the worst
  case, and $O(N \log K)$ time in the best case.
  In practice,
  other implementations can be asymptotically slower, and can 
  return incorrect results. We propose \pkg{binsegRcpp}, an \proglang{R} package which
  provides a correct \proglang{C++} implementation, with the expected asymptotic time complexity.
  %% In practice, existing
  %% implementations such as \pkg{ruptures} (\proglang{Python} module) and
  %% \pkg{changepoint} (\proglang{R} package) are asymptotically slower.
  %% Additionally, \pkg{changepoint} sometimes returns incorrect results.
  %% We present the \proglang{R} package \pkg{binsegRcpp}, which provides an efficient
  %% and correct C++ implementation of binary segmentation.
  We discuss several important \proglang{C++} coding techniques,
  and include detailed comparisons with other implementations of binary
  segmentation: \pkg{ruptures}, \pkg{fpop}, \pkg{changepoint}, \pkg{blockcpd}, and
  \pkg{wbs}.
}

@@

TODO \code{fpop::multiBinSeg} \citep{Maidstone2017}.

TODO \code{wbs::sbs} \citep{Baranowski2019}.

@@

\begin{table}[t!]
  \centering
  \small
\begin{tabular}{l|cccccc}
  package & binsegRcpp & changepoint & wbs & fpop & ruptures & blockcpd\\
  function & binseg & cpt.mean & sbs & multiBinSeg & Binseg & fit\_blockcpd\\
%   > ruptures$version$version
% [1] "1.1.6"
% > sapply(c("binsegRcpp", "changepoint", "wbs", "fpop"), function(p)paste(packageVersion(p)))
%  binsegRcpp changepoint         wbs        fpop 
% "2022.3.29"     "2.2.3"       "1.4" "2019.8.26"
% https://github.com/HaotianXu/changepoints does not have binary segmentation.
  version & 2022.3.29 & 2.2.3 & 1.4 & 2019.8.26 & 1.1.9 & 1.0.0\\
\hline
  weights & yes & no & no & no & no & no\\
  max segs & yes & yes & no & yes & yes & yes\\
  dim & one & one & one & multi & multi & multi \\
  correct & yes & no & yes & yes & yes & yes\\
  losses & 5 & 6 & 1 & 1 & 10 & 5\\
  language & C++ & C & C & C++ & Python & C++\\
  % all propose some loss computation.
  storage & heap & arrays & recursion & heap & LRU cache & heap \\
  space & $O(S)$ & $O(S^2)$ & $O(S)$ & $O(S)$ & $O(S)$ & $O(S)$ \\
  cumsum & yes & yes & yes & yes & no & yes\\
  best & $O(N\log N)$ & $O(N^3)$ & $O(N\log N)$ & $O(N\log N)$ & $O(N^{1.4})$ & $O(N\log N)$\\
  worst & $O(N^2)$ & $O(N^3)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$\\
  CV & yes & no & no & no & no & no\\
  params & all & one & no & no & no &one\\
\end{tabular}
\caption{\label{tab:overview}
  Properties (rows) of six binary segmentation libraries (columns).
Properties are weights (can specify observation-specific positive weights for the loss), max segs (can specify maximum number of segments parameter), dim (dimensions of input data), correct (does code return correct model parameters?), losses (number of loss functions implemented), language (programming language to implement loss), storage (technique used to store cost of splitting a segment), space (space complexity in number of segments $S$), cumsum (cumulative sum used), best (time complexity in best case data), worst (time complexity in worst case data), CV (support for cross-validation), params (parameters are returned). }
\end{table}

>>>>

%% - \Abstract{} almost as usual
\Abstract{
  Binary segmentation is a classic algorithm for detecting
  change-points in sequential data. In theory, using a simple loss
  function like Normal or Poisson, binary segmentation should be
  extremely fast for $N$ data and $K$ segments: asymptotically $O(N K)$ time in the worst
  case, and $O(N \log K)$ time in the best case.
  In practice,
  existing implementations can be asymptotically slower, and can 
  return incorrect results.
  We propose \pkg{binsegRcpp}, an \proglang{R} package which
  provides a correct \proglang{C++} implementation,
  with the expected asymptotic time complexity.
  %% In practice, existing
  %% implementations such as \pkg{ruptures} (\proglang{Python} module) and
  %% \pkg{changepoint} (\proglang{R} package) are asymptotically slower.
  %% Additionally, \pkg{changepoint} sometimes returns incorrect results.
  %% We present the \proglang{R} package \pkg{binsegRcpp}, which provides an efficient
  %% and correct C++ implementation of binary segmentation.
  We discuss several important \proglang{C++} coding techniques,
  and include detailed comparisons with other implementations of binary
  segmentation:
  \pkg{ruptures}, \pkg{fpop}, \pkg{changepoint}, \pkg{blockcpd}, and
  \pkg{wbs}.
}

@@

Change-point detection is an important problem in sequential data
analysis, that occurs in a variety of different applications, such as
genomics \citep{Hocking2013bioinformatics}, neuroscience
\citep{Jewell2019biostatistics}, and medical monitoring
\citep{Fotoohinasab2020segmentationEMBC}, among others.
The focus of
this article is change-point detection in the off-line setting, in
which we assume there is a sequence of $N$ observed data, and we want
an algorithm to identify the precise locations of any abrupt changes
that may be present \citep{Truong2020}. We assume there is
user-specified maximum number of change-points $K$, and we would like
to compute a sequence of models from 1 to $K$ change-points.

Using
the classical dynamic programming algorithm of
\citet{segment-neighborhood}, optimal change-points for a given cost
function, and a sequence of model sizes $k\in\{1,\dots,K\}$, can be
computed in $O(N^2 K)$ time.
%\citep{Jackson2005}
More recently, the pruned dynamic programming algorithms of
\citet{Maidstone2017} can be used to compute the same sequence of $K$
optimal models, with the same $O(N^2 K)$ worst case time complexity,
but faster $O(KN)$ time best case time complexity.  Other dynamic
programming algorithms such as PELT \citep{pelt} and DuST
\citep{truong2024efficient} compute a single model in best case linear
time $O(N)$, worst case quadratic time $O(N^2)$.  They may be used as
a sub-routine to efficiently compute a range of $K$ change-point
models, using the CROCS algorithm \citep{Liehrmann2021chipseq}, which
again results in best case $O(N K)$ time, worst case $O(N^2 K)$ time.

Binary segmentation is a heuristic algorithm for computing a sequence of models from 1 to $K$ change-points \citep{binary-segmentation}.
Using a simple loss function like the square loss, the best case of binary segmentation is $O(N \log K)$ time \citep{Hocking2024finite}, which is much faster than dynamic programming.
The worst case of binary segmentation is $O(N K)$ time, which is the same speed as the best case of dynamic programming.

\begin{table}[t!]
  \centering
  \small
\begin{tabular}{l|cccccc}
%   > ruptures$version$version
% [1] "1.1.6"
% > sapply(c("binsegRcpp", "changepoint", "wbs", "fpop"), function(p)paste(packageVersion(p)))
%  binsegRcpp changepoint         wbs        fpop 
% "2022.3.29"     "2.2.3"       "1.4" "2019.8.26"
% https://github.com/HaotianXu/changepoints does not have binary segmentation.
  package & binsegRcpp & changepoint & wbs & fpop & ruptures & blockcpd\\
  function & binseg & cpt.mean & sbs & multiBinSeg & Binseg & fit\_blockcpd\\
  version & 2022.3.29 & 2.2.3 & 1.4 & 2019.8.26 & 1.1.9 & 1.0.0\\
\hline
weights & yes & no & no & no & no & no\\
min len & yes & yes & no & no & yes & yes\\
max segs & yes & yes & no & yes & yes & yes\\
dim & one & one & one & multi & multi & multi \\
correct & yes & no & yes & yes & yes & yes\\
losses & 5 & 6 & 1 & 1 & 10 & 5\\
language & C++ & C & C & C++ & Python & C++\\
% all propose some loss computation.
storage & heap & arrays & recursion & heap & LRU cache & heap \\
space & $O(S)$ & $O(S^2)$ & $O(S)$ & $O(S)$ & $O(S)$ & $O(S)$ \\
cumsum & yes & yes & yes & yes & no & yes\\
best & $O(N\log N)$ & $O(N^3)$ & $O(N\log N)$ & $O(N\log N)$ & $O(N^{1.4})$ & $O(N\log N)$\\
worst & $O(N^2)$ & $O(N^3)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$ & $O(N^2)$\\
CV & yes & no & no & no & no & no\\
params & all & one & no & no & no &one\\
\end{tabular}
\caption{\label{tab:packages}
  Properties (rows) of six binary segmentation libraries (columns).
Properties are weights (can specify observation-specific positive weights for the loss), min len (can specify minimum segment length parameter), max segs (can specify maximum number of segments parameter), dim (dimensions of input data), correct (does code return correct model parameters?), losses (number of loss functions implemented), language (programming language to implement loss), storage (technique used to store cost of splitting a segment), space (space complexity in number of segments $S$), cumsum (cumulative sum used), best (time complexity in best case data), worst (time complexity in worst case data), CV (support for cross-validation), params (parameters returned, one or all model sizes). }
\end{table}

There are several free/open-source implementations of binary
segmentation, most of which use C/C++ code for efficiency
(Table~\ref{tab:packages}).  The \pkg{fpop} package in R contains the
\code{fpop::multiBinSeg} function, which implements the square loss
for multi-dimensional data \citep{Maidstone2017}.  The \pkg{wbs}
package in R contains the \code{wbs::sbs} function, which implements
the square loss for univariate data \citep{Baranowski2019}.  Since it
uses recursive function calls, it does not support the max segments
parameter $K$, and instead always computes the full path of models
(from 1 to $N-1$ change-points).

@@
